Reconstructing the evolutionary history of freshwater fishes (Nemacheilidae) across Eurasia since early Eocene

Version 1

Abstract

Abstract

Eurasia has undergone substantial tectonic, geological, and climatic changes throughout the Cenozoic era, primarily associated with tectonic plate collisions and a global cooling trend. The evolution of present-day biodiversity unfolded in this dynamic environment, characterised by intricate interactions of abiotic factors. However, comprehensive, large-scale reconstructions illustrating the extent of these influences are lacking. We reconstructed the evolutionary history of the freshwater fish family Nemacheilidae across Eurasia and spanning most of the Cenozoic era. Molecular phylogeny uncovered six major clades within the family, along with numerous unresolved taxonomic issues. Dating of cladogenetic events and ancestral range estimation traced the origin of Nemacheilidae to Indochina around 48 million years ago. Subsequently, one branch of Nemacheilidae colonised eastern, central, and northern Asia, as well as Europe, while another branch expanded into the Burmese region, the Indian subcontinent, the Near East, and northeast Africa. These expansions were facilitated by tectonic connections, favourable climatic conditions, and orogenic processes. Conversely, aridification emerged as the primary cause of extinction events. Our study marks the first comprehensive reconstruction of the evolution of Eurasian freshwater biodiversity on a continental scale and across deep geological time.

Introduction

The present-day species- and genus-level diversity on earth is mainly the result of the evolution during Cenozoic, when biodiversity recovered from the Cretaceous-Tertiary mass extinction event, which eradicated three-quarters of all animal and plant species. During the Cenozoic new lifeforms flourished; often from lineages that had played only a minor role before the mass extinction; like mammals, birds and teleost fishes (Feduccia, 1995; Friedman, 2010).

The Cenozoic is also a period of significant and relatively rapid climatic and geological changes (Summerhayes, 2015). The biggest geological changes occurred in Eurasia, resulting from its collision with the African, Arabian and Indian plates, that caused the closure of the Tethys Sea and the Alpine - Himalayan orogeny (Trifonov et al., 2012). The climate also changed dramatically during Cenozoic; transitioning Earth from a warm and humid greenhouse state to a cool and dry icehouse; but with a complex interplay of large-scale changes and smaller-scale fluctuations, creating a diverse range of climatic conditions on the local scale (Mudelsee et al., 2014).

Geologic and climatic changes are major drivers of biodiversity evolution (Antonelli et al., 2018; Mayhew et al., 2012; Fritz et al., 2016). Numerous biogeographical studies show distribution patterns and local vicariance and radiation events in the Eurasian biodiversity (e.g. Andreyenkova et al., 2021; Ishii et al., 2023). The majority of these studies focused on terrestrial fauna and flora, but we believe that the best suited model organisms for biogeographical analysis are strict freshwater species. Such hololimnic species are confined to a certain water body unless a physical connection with another water body appears. At the same time, the evolution of the earth’s hydrographic network bases on local geological and climatic conditions. The locally ‘locked’ freshwater fauna thus reflects the environmental past of any region on earth in much more detail than most terrestrial animals or plants (e.g. Bănărescu, 1992; Bohlen et al., 2020a).

Despite this potential, there is a paucity of research on the evolutionary history of Eurasian freshwater biodiversity covering the large geographic and temporal scales. Numerous studies deal with certain geologic or climatic changes during short parts of the Cenozoic (e.g. Kahlke 2014; Ballarin & Li 2018; Gu et al. 2022) or cover a comparably small region of Eurasia (e.g. Delić et al. 2020; Deng et al. 2020; Saito et al. 2018), but there is no comprehensive, overarching analysis that covers the whole of Eurasia, most of the Cenozoic and considers climatic as well as geologic events as driving forces for the evolution of biodiversity.

For the present study we chose the species-rich (> 700 species) freshwater fish family Nemacheilidae (stone loaches) as model; which is omnipresent in almost all water bodies of Europe and Asia (Kottelat, 2012; Dvořák et al., 2022a) and stretches from Spain to Japan in west-east direction and from Russia to indonesia in north-south direction. Most species are small (< 12 cm SL), benthic and inhabit streams and rivers, although some are found in swamps, caves and lakes. Few even occupy extreme environments: Nemacheilidae include the deepest freshwater cave fish in the world (Triplophysa gejiuensis, 400 m below surface) and the fish living at the highest altitude (Triplophysa stolickai, 5200 m a.s.l.) (Chu & Chen 1979, Kottelat 2012).

The aim of the study is to reconstruct the evolutionary history of the family Nemacheilidae as example for the evolution of freshwater fauna across Eurasia and since early Cenozoic. The research concept includes a phylogenetic reconstruction based on molecular genetic data, followed by the dating of cladogenetic events and a biogeographic analysis in order to reconstruct ancestral distribution ranges. The results are compared with the known geological and climatic history of Eurasia to identify the causative factors and processes that have influenced the diversification of Eurasian freshwater fishes.

Results

Major clades within Nemacheilidae

To understand the phylogenetic diversity of the family Nemacheilidae and to identify the largest units within, we analysed sequence data for one mitochondrial (cytochrome b) and up to five nuclear (EGR3, IRBP2, Myosin6, RAG1, Rhodopsin1) genes of 471 specimens representing 279 species and 37 genera from across Eurasia (Fig. 1; Supplementary Material Table S1). This is by far the most comprehensive genetic dataset of Nemacheilidae ever published. The outgroup included 21 species from 9 related families (Fig. 2). We applied maximum likelihood analysis (ML) in IQ-TREE and Bayesian inference (BI) in mrBayes on the concatenated dataset. Besides, the individual unrooted ML gene trees were reconstructed in IQ-TREE and subsequently used as input files for ASTRAL III (Zhang et al., 2018) to infer species tree. All analyses revealed Nemacheilidae as monophyletic group and recovered the presence of six monophyletic major clades within the dataset.

Phylogenetic tree of Nemacheilidae, including all 471 analysed nemacheilid specimens, reconstructed in MrBayes based on six loci. The topologies of BI and ML trees were congruent. The values at the basal nodes correspond to Bayesian posterior probabilities and ML bootstrap supports, respectively. Only the ingroup is shown.

Divergence time estimation resulting from Bayesian divergence time analysis of concatenated dataset in BEAST 2: maximum clade credibility (MCC) tree of the whole dataset, with the Nemacheilidae ingroup collapsed into main clades. Red stars indicate calibration points, values at branches represent posterior probabilities (PPs lower than 0.90 not shown) and blue horizontal bars denote 95% HPD. The time scale is in millions of years.

Each major clade exhibited a distinct distribution (Fig. 3). Overlap of ranges occur in the Burmese region and in Indochina.

Geographic distribution of the six major clades of Nemacheilidae.

The most widely distributed is the ‘Northern Clade’ (Fig. 3A), which inhabits the northern parts of Asia and most of Europe from Japan to Spain and southwards to Yunnan. The ‘Southern Clade’ (Fig. 3A) has the second largest distribution and inhabits the Indian subcontinent, the Burmese region and West Asia from the Salween River to southeast Europe; isolated species are found also in Indochina and East Africa (Ethiopia). Another clade is the ‘Burmese Clade’ (Fig. 3B) with a continuous distribution from western Thailand to Northeast India, plus isolated species in the Western Ghats of India and in eastern Thailand and Cambodia. The ‘Sundaic Clade’ (Fig. 3B) includes only the species of the genus Nemacheilus and is distributed across Sundaland (the Great Sunda Islands Borneo, Sumatra and Java and the Malay Peninsula) and in the Indochinese rivers draining to the Gulf of Thailand (Mekong, Mae Khlong and Chao Phraya). The ‘Eastern Clade’ (Fig. 3C) is found along the eastern margin of Asia from the Russian Far East to eastern Borneo. Its disjunct distribution consists of a single species in northeast Borneo, the majority of genera and species in northern Vietnam and southeast China and a single genus on the Korean Peninsula, the Japanese Archipelago, northeast China and the Russian Far East.

The sixth clade (‘Indochinese Clade’, Fig. 3C) has a continuous distribution throughout Thailand, Laos, Cambodia, Vietnam, central Myanmar and southern China, as well as isolated species in the upper/middle Yangtze and on Tioman Island (Malaysia). The distribution areas of the six major clades are not exclusive, but show little overlap in the cases of Northern, Indochinese and Southern Clades, while the Sundaic Clade shares the northern half of its distribution area with the Indochinese Clade and the Eastern Clade even shares its northern part of distribution with the Northern Clade, the middle part with the Indochinese Clade and the southernmost point of occurrence with the Sundaic Clade.

Analyses of the concatenated dataset as well as ASTRAL analysis inferring a species tree from single gene trees revealed a basal dichotomy in the family Nemacheilidae (Fig. 1,2,4). The one branch consists of the Eastern Clade as sister to the pair Indochinese plus Northern Clade. The second branch is made of the Sundaic Clade as sister to the pair Burmese and Southern Clade. All basal nodes and most of tip nodes are supported by high statistical supports in all analyses.

Topologies of main clades derived from various analyses. A) Topologies recovered by ML analyses of single gene trees, with values at branches indicating bootstrap supports. Topologies congruent to the concatenated data-derived as well as species tree is revealed by Cyt b, MYH6 and IRBP2. Three slightly different topologies were derived from RH, EGR3 and RAG1 B) Topology obtained from analyses of concatenate dataset, the values at the branches correspond to ML/MrBayes PP/BEAST PP supports, respectively. C) Topology inferred by ASTRAL-III using the unrooted ML single gene trees as input. The values at the branches correspond to support values/gCF/sCF.

Divergence times

The ages of the genealogic events were determined using three fossil records and one geological event as calibration points. The results indicate that Nemacheilidae shared the last common ancestor with other families of Cobitoidea about 48.5 mya (Fig. 4). The first notable split that separated the family into two big branches took place around 39 mya (for details see Fig. 5).

Divergence time estimation: Ingroup of the maximum clade credibility (MCC) tree resulting from Bayesian divergence time analysis of concatenated dataset in BEAST 2. The red star indicates internal calibration point, the values at the nodes represent posterior probabilities (PPs lower than 0.90 not shown), and the blue bars relevant 95% HPD. The time scale is in millions of years. Inset in top left corner shows complete tree: Nemacheilidae is highlighted in yellow, while the outgroup is not highlighted; all calibration points are indicated by red stars.

Biogeographic reconstructions

The biogeographic reconstruction of Nemacheilidae considered 11 geographic regions within Eurasia and northeast Africa. The results show that Nemacheilidae originated in the region of nowadays Indochina and expanded from there in multiple waves. More details are given in Fig. 6.

Reconstruction of the biogeographical history of Nemacheilidae. The inset map in top left corner shows the defined biogeographical regions of Eurasia recognised in the analysis; below the map the colour – region code for the main tree. Grey colour indicates combination of two regions as ancestral (FH); black colour indicates ancestral ranges identified with likelihood < 10%. Main tree: Ancestral range reconstruction with use of DEC+J model based on the phylogeny derived from BEAST2. Pie charts at the nodes indicate relative likelihood of ancestral ranges of most recent common ancestors (MRCA), while the current distribution of taxa is indicated at the tips.

Discussion

The evolutionary history of Nemacheilidae (Fig. 7)

Early Eocene (56 – 48 mya) - the origin of Nemacheilidae (Fig. 8A)

The family Nemacheilidae shared the last common ancestor with the other Cypriniformes about 48.5 mya (Fig. 2), and our biogeographic analysis identified the region of nowadays Indochina as most likely place of origin of Nemacheilidae (Fig. 6). During the early Eocene, Southeast Asia experienced a warm and perhumid climate, characterised by both highlands and lowlands with numerous rivers (Morley 2018), which must have been ideal conditions for stone loaches.

Time tree of the evolutionary history of Nemacheilidae: The figure presents the evolutionary history of Nemacheildae, highlighting the dating of key geological or climatic key paleo-events that might have significantly impacted their evolution. Orange bars represent exceptionally warm periods, blue bars denote cold periods. Simple boxes give time span of specific geological or climatic event. For detailed explanations of these events and their impact on nemacheilid evolution refer to text.

Series of paleomaps visualising the geographical and chronological aspects of the main events during the evolutionary history of the freshwater fish family Nemacheilidae across Eurasia. A red star marks the point of origin of Nemacheilidae, arrows indicate dispersal events.

Middle to late Eocene (48 – 34 mya) – first range expansions and formation of the first major clades (Fig. 8B,C)

The first detectable cladogenetic event within the family occurred about 39 mya and separated a lineage formed by the Eastern, Northern and Indochinese Clades (Branch A in Fig. 5) from a lineage including the Sundaic, Burmese and Southern Clades (Branch B). Our biogeographic analysis of the dataset revealed Indochina as most parsimony area of the first cladogenetic event; and consequently as region of origin for branch A as well as branch B.

The Eastern Clade was the first of the six present major clades to form: it separated about 35.8 mya from the common ancestor of the Indochinese and Northern Clades, either in the region of Indochina or Vietnam-south China. Taking into consideration its present-day distribution (Fig. 3) we assume that the young Eastern Clade spread mainly along the eastern coastline of Asia southwards to Borneo and northwards to northeast China. Geologic data support this hypothesis: During the Eocene, Asia was much smaller than today and had a rather flat relief with old mountains mainly along its eastern coast due to the collision with the Pacific Plate since 190 mya (Seton et al., 2012; Wang, 2004), while the upfolding of east-west mountain ridges had just started. The generally warm and humid climate during Eocene (Morley 2018; Westerhold et al. 2020) promoted the range expansion of the Eastern Clade, while the ancient eastern mountain ranges directed the range expansion. It can be assumed that shortly after the range expansion the distribution of the Eastern Clade was more or less continuous and not separated into several disjunct areas as nowadays.

The ancestral Indochinese + Northern Clades also spread northwards, but their present-day distribution (Fig. 3) suggests that they took a route through Central Indochina and China. Most likely, this first geographic expansion led the ancestral Indochinese+Northern Clades northward to at least the Tibetan Protoplateau, which during Eocene epoch had its northern boundary in the region of the Tanggula Shan (Wang et al., 2008). The presence of the two protoclades on the early Tibetan Plateau is indicated by the presence of some of their most basal members in the Yangtze basin and Yunnan: the genus Homatula for the Indochinese Clade and several species of Triplophysa for the Northern Clade (Figs. 1,5).

Oligocene (34 – 23 mya) – range expansions, extinctions and the formation of the remaining major clades (Fig. 8D,E)

The Eocene - Oligocene transition brought a period of significant global cooling (Hutchinson et al 2021) as well as a change in the precipitation regime of Central Asia (Clift & Webb 2018; Ye et al. 2022) due to the ongoing uplift of the Himalayan Mountains (Ding et al. 2017). A substantial decrease of rainfall in Central Asia caused an aridification from 34 – 23 mya (Bosboom et al. 2014) extending across nearly all of Oligocene Asia from the western to the eastern coast (Wang, 2004). The distribution area of the Eastern Clade became fragmented during this period (Fig. 3).

The Central Asian arid zone also split the Indochinese – Northern protoclade into two part that evolved into two separate major clades, the Indochinese Clade and the Northern Clade. Moreover, the distribution area of the Indochinese Clade was split and separated the southernmost branch (Speonectes and Tuberoschistura) from the northern branch. The latter presently inhabits all of Indochina and southern China including the middle Yangtze basin (Fig. 3), indicating that the Oligocene refuge of the northern branch of the Indochinese Clade was located southeast of the Tibetan Plateau. In contrast, the early Northern Clade found its refuge further north on the Tibetan Protoplateau, the only region that can explain the further history and the present distribution of the clade. By late Oligocene, the climate in Central Asia returned to a warmer and moister state (Zachos et al. 2001) and the eastern Kunlun Shan Mountains rapidly thrust upward as an expansion of the northern Tibetan Plateau (Li et al. 2021). These favourable conditions coincided with a major split within the Northern Clade at 20 mya (Fig. 5), when the proto-Barbatula clade colonised the Kunlun Mountains, while the proto Triplophysa branch found its main distribution rather east of the Tibetan Plateau.

Different geologic events drove the evolution of branch B: the common ancestor of the Burmese and Southern Clades split at 32.5 mya in Indochina from the Sundaic Clade. In contrast to the general trend among Nemacheilidae, the species of the Sundaic Clade (genus Nemacheilus) prefer sandy bottom over gravel bottom, making lowlands regions of Sundaland and Indochina their ideal habitat. There are no indications that the Sundaic Clade ever left the region of present-day Southeast Asia.

The separation of the ancestral Burmese and Southern Clades from the Sundaic Clade occurred because the former left Southeast Asia by colonising the West Burma Terrane, a microplate that was pushed northward by the Indian plate (Morley et al. 2021) and made temporary contact with Sundaland around Eocene-early Oligocene (Westerweel et al. 2019). Around 30 mya, it occasionally made contact with northeast India (Morley et al. 2021), allowing fishes from the West Burma Terrane to colonise the Indian subcontinent. During the Oligocene, the Indian plate and the West Burma Terrane both collided with mainland Asia. The former presently forms the Indian subcontinent, the later Western Burma (Westerweel et al. 2020). Their parallel collision with mainland Asia caused the uplift of the north-south directed Indo-Burmese mountain range (Morley et al. 2020), providing a physical barrier between the Brahmaputra basin on the Indian side from the Chindwin-Irrawaddy basin on the Burmese side, promoting the separation of the two faunas into the Burmese Clade and the Southern Clade.

Miocene (23 – 5.3 mya) – rapid diversification (Fig. 8F,G)

The Miocene is generally an epoch of cooling, with temperatures decreasing about 5.5° C to conditions similar to present day (Zachos et al. 2001). But temperatures also underwent strong fluctuations, leading to exceptionally warm periods like the Middle Miocene Climatic Optimum (17 – 14.5 mya) as well as to drastic cooling events like the Middle Miocene Climatic transition (14 mya) (Morley 2018). Additionally, the Asian monsoon regime strengthened from the early Miocene (Clift & Webb 2018), increasing precipitation and altering the distribution of climatic regions across Asia (Sun & Wang 2005). Moreover, during the Miocene the Himalayan Mountains experienced their main uplift (Ding et al. 2017) and the upfolding area of Central Asia mountain ranges expanded northwards, giving rise to the Tian Shan and Altai mountains (Nissen et al 2009; Liu et al 2023).

During this complex geological and climatic scenario, the fish family Nemacheilidae underwent a massive diversification. In our phylogenetic hypothesis the family comprised 16 lineages at the beginning of Miocene and 177 lineages by the end of the epoch. The radiation occurred across all six major clades, albeit to a lesser extent in the Eastern Clade. Interestingly, it manifested similarly in the region of the Indian monsoon and the East Asian monsoon. In both regions, major radiations occurred during the Middle Miocene Climatic Optimum, including the radiation of the Barbatula lineage in the Northern Clade, of the Indochinese Clade, of all Nemacheilus except N. binotatus and N. ornatus and of six branches in the Southern Clade. It appears that under increased temperature and precipitation freshwater fishes flourished, across all of Asia and under different monsoon systems. One of the radiations that began during the Middle Miocene Climate Optimum involves Afronemacheilus, the sole nemacheilid genus found in Africa (Prokofiev & Golubtsov 2013). Our results confirm that the ancestral Afronemacheilus separated from the Southern Clade about 18 - 17 mya. It could have utilised two potential pathways to Africa: one across the Arabic Peninsula and via the southern Danakil Isthmus, a broad landbridge that stretched from Djibouti to Aden from late Miocene to Pliocene (Bosworth et al. 2005; Steward & Murray 2017). The second pathway would lead through the Levant and via Sinai into the Nile (‘Gomphotherium landbridge’ Harzhauser et al 2007), and subsequently upstream into the Ethiopian highlands. Unfortunately, the nearly complete extinction of freshwater fishes from the Arabic Peninsula and from the Nile during Pleistocene aridification (Kotwicki & Salaimani 2009; Leplongeon 2021) complicates the reconstruction of its pathway to Africa.

In the Northern Clade, the Triplophysa lineage flourished during the Miocene. Its present wide distribution in southern and Central China suggests that it expanded from its Oligocene refuge on the eastern Part of the Tibetan Plateau in an eastern direction (Fig. 3). In contrast, the Barbatula branch expanded from its refuge on the western Tibetan Plateau northwards, traversing the uplifting Tian Shan and Altai mountains. This expansion can be traced from the geographic position of the most basal species (T. labiata-Tian Shan, B. altayensis – Altai), and then eastwards (B. compressirostris + D. dzebuadzei – west Mongolia, B. cf. toni – east Mongolia; tip species in Korea (B. indet Korea) and Japan (B. oreas).

During late Miocene, Barbatula also expanded westwards through Siberia and colonised Europe between 5.6 mya (separation of the European taxa from the Mongolian sister taxon) and 3.0 mya (oldest split within European Barbatula). During this cool and dry period, first glaciations occurred on the northern hemisphere (Zachos et al. 2001), and in Siberia the mean temperature of the coldest month dropped below 0° for the first time (Popova et al. 2012). Freezing of the northern margin of Eurasia impeded the Siberian rivers (which flow from south to north) from discharging their water into the Northern Ocean and produce extensive wetlands on the shallow relief of Siberia. These wetlands served as pathways for freshwater fishes from East Asia to European river systems during Pleistocene (Kalous et al. 2012), and we assume that the same mechanism brought Barbatula to Europe during Pliocene.

Pliocene and Pleistocene (5.3 – 0.01 mya) - Ice ages

During Pliocene and Pleistocene global cooling continued until massive glaciations repeatedly covered the north of Eurasia from early Pliocene onwards (de Schepper et al. 2014). However, the Pliocene also brought a much stronger seasonality to Eurasia, especially in the form of cold winters (Shen et al. 2018). Despite these significant changes in climatic conditions, the evolution of Nemacheilidae seems not to have experienced any dramatic negative impact, but reveals mainly local speciation events (e.g. in the genera Traccatichthys and Barbatula). We know from more detailed studies on Nemacheilidae (Bohlen et al 2020a,b; Šlechtová et al. 2021) that during Pliocene and Pleistocene also geographic shifts as well as expansions and disappearances occurred, but since these happened on rather local scale they were too insignificant to reflect in the present intercontinental dataset.

Conclusions

- We reconstruct the evolutionary history of the Eurasian fauna during most of Cenozoic era on the example of the freshwater fish family Nemacheilidae. The extensive dataset (471 specimens, 279 species, 6 genes) from across Eurasia and the low dispersal capacity of the model resulted in high resolution of the phylogenetical and biogeographic analysis.

- Molecular phylogeny uncovered six major clades within the family. Dating of cladogenetic events and ancestral range estimation traced the origin of Nemacheilidae to Indochina around 48 million years ago.

- Major expansions were facilitated by tectonic connections, favourable climatic conditions, and orogenic processes. Conversely, aridification emerged as the primary cause of extinction events.

- Our study represents the first comprehensive and holistic reconstruction of the evolution of Eurasian freshwater biodiversity on a continental scale and through deep geological time.

Methods

Sampling and Laboratory procedures

471 samples of more than 250 species from 37 described genera of the family Nemacheilidae were examined, including the set of sequences of 364 specimens analysed in our laboratory and sequences of 107 specimens obtained from GenBank. Specimens used for the present study were fixed and stored in 96% ethanol. For more details about samples, sampling sites and GenBank accession numbers see Supplementary material Table 1.

Total genomic DNA was isolated using the DNeasy Blood & Tissue kit (Qiagen) following the manufacturer’s instructions. One mitochondrial gene (cytochrome b) and five nuclear genes (RAG1, IRBP2, MYH6, RH1 and EGR3) were selected. For list of primers see Supplementary material Table 2, details about PCRs settings are described in Dvořák et al., 2022; Chen et al., 2003 and Chen et al., 2008.

Sequencing of the PCR products was performed in Laboratory of Fish Genetics on ABI Prism 3130 GA or via sequencing service in Macrogen Europe BV (Amsterdam, Netherlands).

Phylogenetic analyses

Chromatograms were checked and assembled in the SeqMan II module of the DNA Star software package (LASERGENE). Single gene alignments were done in BioEdit (Hall, 1999) with use of ClustalW (Larkin et al., 2007) multiple alignment algorithm. TheThe datasets were concatenated in PhyloSuite v1.2.2 (Zhang et al., 2020).

The best-fit substitution models and partitioning schemes (Supplementary material Table 3) were estimated using Partition Finder 2 (Lanfear et al., 2016) implemented in PhyloSuite 1.2.2 based on the corrected Akaike Information Criterion (AICc).

The 471 specimen phylogenetic trees were inferred from six loci concatenated dataset using the maximum likelihood (ML) and Bayesian inference (BI). The ML analyses were performed using IQ-TREE (Nguyen et al., 2015) implemented in PhyloSuite. The best-fit evolutionary model for each codon partition was automatically determined by ModelFinder (Kalyaanamoorthy, et al. 2017). The node support values were obtained with 5000 ultrafast bootstrap replicates (UFBoot) (Hoang et al., 2018). For the BI analyses we used MrBayes 3.2.7a (Ronquist & Huelsenbeck, 2003) on CIPRES Science Gateway (Miller et al., 2010). The datasets were partitioned into genes and codon positions and analyses were performed in two parallel runs of 10-20 million generations with 8 Metropolis Coupled Markov Chains Monte Carlo (MCMCMC). The relative burnin of 25% was used and from the remaining trees a 50% majority rule consensus trees were constructed.

For reconstructing of the species tree we used the Accurate Species TRee ALgorithm (ASTRAL III) (Zhang et al., 2018). For ASTRAL we used unrooted single gene ML trees reconstructed in IQ-TREE implemented in PhyloSuite. We have used IQ-TREE 2 (Minh at al., 2020) to calculate gene concordance factor (gCF, indicating the percentage of decisive gene trees containing particular branch) and site concordance factor (sCF, indicating the percentage of decisive alignment sites that support a branch in the reference tree) for every branch in the ASTRAL species tree. Distribution maps in Fig. 2 were designed using HYDROSHEDS (Lehner & Grill 2013).

Divergence time estimations and ancestral range reconstruction

The ages of cladogenetic events were estimated in BEAST 2.6.4 (Bouckaert et al., 2014) via CIPRES Science Gateway with use of four calibration points. The first calibration point is based on the oldest known fossil of the family Catostomidae, Wilsonium brevipinne, from early Eocene (Liu, 2021), approximately 56-48 my old. The second calibration point is derived from the only known nemacheilid fossil record, Triplophysa opinata from Kyrgyzstan from middle-upper Miocene (16.0 to 5.3 mya) (Prokofiev, 2007; Böhme and Ilg, 2003). Third calibration point is based on the oldest known fossil of the genus Cobitis, C. naningensis, from early to middle Oligocene in southern China, 34-28 my old (Chen et al., 2015). The fourth calibration point, based on the river history of the southern Korean Peninsula, dates the separation between the species Cobitis tetralineata and C. lutheri to 2.5-3.5 my (Kwan et al. 2014). For the fossil-based calibration points we have used a fossil calibration prior implemented in CladeAge (Matschiner et al., 2017), the fourth calibration point was set to a uniform prior (2.5 – 3.5). The calibration points are depicted in Figs. 2 and 5.

For the final analyses, the partitions were unlinked and assigned the estimated evolutionary models. We used relaxed lognormal molecular clock and birth-death prior. Given the large and complex nature of our dataset, and in light of several preliminary analyses where MCMC chains did not provide satisfactory ESSs even after 1 billion iterations, made the decision to streamline the dataset by retaining only one specimen per species. To enhance the performance of the analysis, we also implemented parallel tempering using the CoupledMCMC package (Müller & Bouckaert, 2019, 2020). The analysis was configured with four parallel chains of 6×108 generations, with resampling every 1000. Tree and parameter sampling intervals were set to 60.000. The effective sampling sizes (ESS) for all parameters were assessed in Tracer 1.7.1 (Rambaut et al., 2018). A maximum clade credibility (MCC) tree was built in TreeAnnotator 2.6.0 (Rambaut & Drummond, 2010) after discarding the first 25% of trees. The final trees were visualised in FigTree 1.4.4 (Rambaut, 2019). ML and BI were performed with both full as well as the reduced datasets.

The biogeographical analysis was conducted using the BioGeoBEARS package (Matzke, 2013) implemented in RASP 4.0 (Reconstruct Ancestral Stage in Phylogenies, Yu et al., 2015). We set 11 biogeographic regions according to drainage areas separated by major mountain ridges (Ural, Caucasus. Himalayas, Tenasserim Ridge, Indo-Burmese Range, Annamite Cordilliera, Balkan, Hindukush) or saltwaser straits (Red Sea) (see Fig. 4A). For the analysis, we utilized the final MCC tree and 100 post burnin trees from the previous BEAST analysis. A maximum of 2-unit areas was allowed to be present in every reconstructed ancestral range. The comparison of six biogeographic models (DEC, DEC+J, DIVALIKE, DIVALIKE+J, BAYAREALIKE, BAYAREALIKE+J) suggested that the DEC+J is the best-suited model for the given dataset (as indicated by AICc weight value, Supplementary material Table 4). Likelihood Ratio Test (LRT) p-values, comparing models with and without the J factor, indicated that the addition of the J factor significantly influences the model likelihood. Considering recent criticisms of DEC and DEC+J (Ree & Sanmartin, 2018), we opted to conduct the analysis not only with DEC+J but also with other +J models. Comparison of the results from all applied models revealed only minimal differences.

eLife Assessment

In this valuable study, the authors attempt to reconstruct the evolutionary history of a large and widespread group of freshwater fishes (Nemacheilidae) across Eurasia since the early Eocene, based on molecular phylogenetic analysis with very comprehensive samplings including 471 specimens belonging to 250 living species. The authors infer that range expansions of the family were facilitated by tectonic connections, favourable climatic conditions, and orogenic processes, adding to our understanding of the effects of climatic change on biodiversity during the Cenozoic. The molecular evidence is overall solid, but the calibration points from the fossil records used in the analysis have not been clearly demonstrated or cited; the different dates for the calibration points might impact the discussion on the evolutionary history relating to past climatic changes.

Reviewer #1 (Public review):

Summary:

This is by far the phylogenetic analysis with the most comprehensive coverage for the Nemacheilidae family in Cobitoidea. It is a much-lauded effort. The conclusions derived using phylogenetic tools coincide with geological events, though not without difficulties (Africa pathway).

Strengths:

Comprehensive use of genetic tools

Weaknesses:

Lack of more fossil records.

Reviewer #2 (Public review):

Summary:

The authors present the results of molecular phylogenetic analysis with very comprehensive samplings including 471 specimens belonging to 250 species, trying to give a holistic reconstruction of the evolutionary history of freshwater fishes (Nemacheilidae) across Eurasia since the early Eocene. This is of great interest to general readers.

Strengths:

They provide very vast data and conduct comprehensive analyses. They suggested that Nemacheilidae contain 6 major clades, and the earliest differentiation can be dated to the early Eocene.

Weaknesses:

The analysis is incomplete, and the manuscript discussion is not well organized. The authors did not discuss the systematic problems that widely exist. They also did not use the conventional way to discuss the evolutionary process of branches or clades, but just chronologically described the overall history.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Summary:

This is by far the phylogenetic analysis with the most comprehensive coverage for the Nemacheilidae family in Cobitoidea. It is a much-lauded effort. The conclusions derived using phylogenetic tools coincide with geological events, though not without difficulties (Africa pathway).

Strengths:

Comprehensive use of genetic tools

Weaknesses:

Lack of more fossil records

Thank you for appreciating the comprehensiveness of our study.

We agree that additional nemacheilid fossils would have provided valuable support for reconstructing the evolutionary history of the family. However, the nemacheilid fossil used in our study is currently the only fossil species of the family, which precludes the possibility of including more. To address this limitation, we incorporated fossils from closely related fish families, as well as a geological event, to calibrate the time tree. We have added further details on this point in “Divergence time estimations and ancestral range reconstruction” section of the Methods. The reconstruction of the pathway by which loaches reached northeast Africa, is further complicated by the extensive aridification of the Arabian Peninsula and the Nile valley, leaving no fossil or extant Nemacheilidae species of Nemacheilidae to provide insights into the distribution of the family during late Miocene.

Reviewer #2 (Public review):

Summary:

The authors present the results of molecular phylogenetic analysis with very comprehensive samplings including 471 specimens belonging to 250 species, trying to give a holistic reconstruction of the evolutionary history of freshwater fishes (Nemacheilidae) across Eurasia since the early Eocene. This is of great interest to general readers.

Strengths:

They provide very vast data and conduct comprehensive analyses. They suggested that Nemacheilidae contain 6 major clades, and the earliest differentiation can be dated to the early Eocene.

Weaknesses:

The analysis is incomplete, and the manuscript discussion is not well organized. The authors did not discuss the systematic problems that widely exist. They also did not use the conventional way to discuss the evolutionary process of branches or clades, but just chronologically described the overall history.

In the revised version, we address the systematic issues within Nemacheilidae in a new paragraph. The polyphyly of the genus Schistura and the polyphyly or paraphyly of many other nemacheilid genera are wellknown challenges in ichthyology. However, the large size of the family Nemacheilidae and the absence of a clear basal classification system has made systematic work difficult.

The chronological concept in the description of events is in accordance with the sequence in which the events occurred over time and corresponds with Figure 8. Additionally, a clade-by-clade description would make it challenging to capture the periods before all clades were formed. As a compromise, the revised version includes a new table where each clade is represented by a column, allowing readers to trace the history of each clade in a clear overview. With this table, we make both the chronological and clade-by-clade perspectives to enhance reader understanding

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

I have no major comments, except for Figure 8, where the colour code for Sunda is not consistent, appearing as light purple and then dark purple. I was trying to locate the colour legend, maybe include this for all figures or refer to it.

Figure 8 has been revised to improve matching of the colours.

Reviewer #2 (Recommendations for the authors):

(1) It is better to discuss the evolutionary history of the major inner groups. For example, why the Branch A and B differentiated? How are the 6 major clades differentiated?

As mentioned above, the new table provides an overview of the evolutionary history of the major clades and, where known, the mechanism that led to their differentiation. For branches A and B, the underlying causes of differentiation remain known. Currently, the extensive morphological variability within each clade prevents a definitive morphological diagnosis, but such a study is planned for the future.

(2) In this study, there are still some phylogenetic or systematic problems unresolved. For example, the Genus Schistura remains polyphyletic even in different major clades. The situation is similar for the Genus Tripophysa though not so serious. These need to be discussed or at least partially solved before discussing the evolutionary history.

We discuss these topics now in a new paragraph ‘Taxonomic implications’.

(3) In Table S1, what is the meaning of "-". Does this mean no data available? If so, how do the authors treat this in their phylogenetic analysis?

Indeed, in Table S1, a ‘-‘ indicates that no sequence was available for the given species and gene. In the phylogenetic analyses, these cases were treated as missing data.

(4) What is the source of Figure 8? There are different opinions on the geological events. The authors need to indicate the source of their information.

The sources of Fig. 8 are now provided in the figure caption.

(5) The Eastern Clade forms continuous distribution in Figure 6, but discontinuous in Figure 8. Is this correct?

Figure 6 does not display the distribution areas for the clades, but illustrates the biogeographic regions used in the biogeographic analysis.

Version 2

Abstract

Abstract

Eurasia has undergone substantial tectonic, geological, and climatic changes throughout the Cenozoic era, primarily associated with tectonic plate collisions and a global cooling trend. The evolution of present-day biodiversity unfolded in this dynamic environment, characterised by intricate interactions of abiotic factors. However, comprehensive, large-scale reconstructions illustrating the extent of these influences are lacking. We reconstructed the evolutionary history of the freshwater fish family Nemacheilidae across Eurasia and spanning most of the Cenozoic era on the base of 471 specimens representing 279 species and 37 genera. Molecular phylogeny using 6 genes uncovered six major clades within the family, along with numerous unresolved taxonomic issues. Dating of cladogenetic events and ancestral range estimation traced the origin of Nemacheilidae to Indochina around 48 million years ago. Subsequently, one branch of Nemacheilidae colonised eastern, central, and northern Asia, as well as Europe, while another branch expanded into the Burmese region, the Indian subcontinent, the Near East, and northeast Africa. These expansions were facilitated by tectonic connections, favourable climatic conditions, and orogenic processes. Conversely, aridification emerged as the primary cause of extinction events. Our study marks the first comprehensive reconstruction of the evolution of Eurasian freshwater biodiversity on a continental scale and across deep geological time.

Introduction

The present-day species- and genus-level diversity on earth is mainly the result of the evolution during Cenozoic, when biodiversity recovered from the Cretaceous-Tertiary mass extinction event, which eradicated three-quarters of all animal and plant species. During the Cenozoic new lifeforms flourished; often from lineages that had played only a minor role before the mass extinction; like mammals, birds and teleost fishes (Feduccia, 1995; Friedman, 2010).

The Cenozoic is also a period of significant and relatively rapid climatic and geological changes (Summerhayes, 2015). The biggest geological changes occurred in Eurasia, resulting from its collision with the African, Arabian and Indian plates, that caused the closure of the Tethys Sea and the Alpine - Himalayan orogeny (Trifonov et al., 2012). The climate also changed dramatically during Cenozoic; transitioning Earth from a warm and humid hothouse state to a cool and dry icehouse (Westerhald et al., 2020); but with a complex interplay of large-scale changes and smaller-scale fluctuations, creating a diverse range of climatic conditions on the local scale (Mudelsee et al., 2014).

Geological and climatic changes are major drivers of biodiversity evolution (Antonelli et al., 2018; Fritz et al., 2016; Mayhew et al., 2012). Numerous biogeographical studies show distribution patterns and local vicariance and radiation events in the Eurasian biodiversity (e.g. Andreyenkova et al., 2021; Ishii et al., 2023). The majority of these studies focused on terrestrial fauna and flora, but we believe that the best suited model organisms for biogeographical analysis are strict freshwater species. Such hololimnic species are confined to a given water body unless a physical connection with another water body appears. At the same time, the evolution of the earth’s hydrographic network is caused by local geomophological and climatic conditions. The locally ‘locked’ freshwater fauna thus reflects the environmental past of any region on earth in much more detail than most terrestrial animals or plants (e.g. Bănărescu, 1992; Bohlen et al., 2020a).

Despite this potential, there is a paucity of research on the evolutionary history of Eurasian freshwater biodiversity covering the large geographic and temporal scales. While numerous studies have examined certain geological or climatic changes during specific periods of the Cenozoic (e.g. Ballarin and Li, 2018; Gu et al., 2022; Kahlke, 2014) or within limited regions of Eurasia (e.g. Delić et al., 2020; Deng et al., 2020; Saito et al., 2018), a comprehensive analysis encompassing the whole Eurasia and through most of the Cenozoic era and considering climatic as well as geological events as driving forces of biodiversity evolution is lacking.

For the present study we chose the species-rich (presently 838 species) freshwater fish family Nemacheilidae (stone loaches) as model; which is present in almost all water bodies of Europe and Asia (Dvořák et al., 2022; Kottelat, 2012) and stretches from Spain to Japan in west-east direction and from Russia to Indonesia in north-south direction. Most species are small (< 12 cm SL), benthic and inhabit streams and rivers, although a few are found in swamps, caves and lakes. Few even occupy extreme environments: Nemacheilidae include the deepest freshwater cave fish in the world (Triplophysa gejiuensis, 400 m below surface) and the fish living at the highest altitude (Triplophysa stolickai, 5200 m a.s.l.) (Chu and Chen, 1979; Kottelat, 2012).

The aim of this study is to reconstruct the evolutionary history of the family Nemacheilidae as example for the evolution of freshwater fauna across Eurasia and since early Cenozoic. The research concept includes a phylogenetic reconstruction based on molecular genetic data, followed by the dating of cladogenetic events and a biogeographic analysis in order to reconstruct ancestral distribution ranges. The results are compared with the known geological and climatic history of Eurasia to identify the causative factors and processes that have influenced the diversification of Eurasian freshwater fishes.

Results

Major clades within Nemacheilidae

To understand the phylogenetic diversity of the family Nemacheilidae and to identify the largest units within, we analysed sequence data for one mitochondrial (cytochrome b) and up to five nuclear (EGR3, IRBP2, Myosin6, RAG1, Rhodopsin1) genes of 471 specimens representing 279 species and 37 genera from across Eurasia (Fig. 1; Supplementary Material Table S1). This is by far the most comprehensive genetic dataset of Nemacheilidae ever published. The outgroup included 21 species from 9 related families (Fig. 2). We applied maximum likelihood analysis (ML) in IQ-TREE and Bayesian inference (BI) in mrBayes on the concatenated dataset. Besides, the individual unrooted ML gene trees were reconstructed in IQ-TREE and subsequently used as input files for ASTRAL III (Zhang et al., 2018) to infer species tree. All analyses revealed Nemacheilidae as monophyletic group and recovered the presence of six monophyletic major clades within the dataset.

Phylogenetic tree of Nemacheilidae, including all 471 analysed nemacheilid specimens, reconstructed in MrBayes based on six loci.

The topologies of BI and ML trees were congruent. The values at the basal nodes correspond to Bayesian posterior probabilities and ML bootstrap supports, respectively. Only the ingroup is shown.

Divergence time estimation resulting from Bayesian divergence time analysis of concatenated dataset in BEAST 2: maximum clade credibility (MCC) tree of the whole dataset, with the Nemacheilidae ingroup collapsed into main clades.

Red stars indicate calibration points, values at branches represent posterior probabilities (PPs lower than 0.90 not shown) and blue horizontal bars denote 95% HPD. The time scale is in millions of years.

Each major clade exhibited a distinct distribution (Fig. 3). The most widely distributed is the ‘Northern Clade’ (Fig. 3A), which inhabits the northern parts of Asia and most of Europe from Japan to Spain and southwards to Yunnan. In our analyses it includes all species of the genera Barbatula, Claea and Triplophysa. The ‘Southern Clade’ (Fig. 3A) has the second largest distribution and inhabits the Indian subcontinent, the Burmese region and West Asia from the Salween River to southeast Europe; isolated species are found also in Indochina and East Africa (Ethiopia). The Southern Clade is the clade with the highest taxonomic diversity including all analysed species of 14 genera (Afronemacheilus, Mesonoemacheilus, Mustura, Nemachilichthys, Neonoemacheilus, Oxynoemacheilus, Paracobitis, Paraschistura, Petruichthys, Physoschistura, Pteronemacheilus, Sasanidus, Seminoemacheilus and Turcinoemacheilus) plus several species of the genus Schistura. Another clade is the ‘Burmese Clade’ (Fig. 3B) with a continuous distribution from western Thailand to Northeast India, plus isolated species in the Western Ghats of India and in eastern Thailand and Cambodia. It comprises all analysed species of the genera Aborichthys, Acanthocobitis and Paracanthocobitis, as well as several species classified as Schistura. The ‘Sundaic Clade’ (Fig. 3B) includes only the species of the genus Nemacheilus and is distributed across Sundaland (the Great Sunda Islands Borneo, Sumatra and Java and the Malay Peninsula) and in the Indochinese rivers draining to the Gulf of Thailand (Mekong, Mae Khlong and Chao Phraya). The ‘Eastern Clade’ (Fig. 3C) is found along the eastern margin of Asia from the Russian Far East to eastern Borneo. Its disjunct distribution consists of a single genus (Sundoreonectes) in northeast Borneo, the majority of genera and species in northern Vietnam and southeast China and a single genus (Lefua) on the Korean Peninsula, the Japanese Archipelago, northeast China and the Russian Far East. The Eastern Clade comprises all analysed species of the genera Karstsinnectes, Lefua, Micronemacheilus, Oreonectes, Paranemachilus, Sundoreonectes, Traccatichthys, Troglonectes and Yunnanilus. The sixth clade (‘Indochinese Clade’, Fig. 3C) has a continuous distribution throughout Thailand, Laos, Cambodia, Vietnam, central Myanmar and southern China, as well as isolated species in the upper/middle Yangtze and on Tioman Island (Malaysia). It is composed of all analysed species belonging to Homatula, Rhyacoschistura, Sectoria, Spaeonectes and Tuberoschistura, in addition to many species currently assigned to Schistura and four species of uncertain generic assignment (‘Nemacheilus’).

Geographic distribution of the six major clades of Nemacheilidae.

The distribution areas of the six major clades are not mutually exclusive. While the Northern, Indochinese, and Southern Clades exhibit limited overlap with each other, the Sundaic Clade shares its northern range with the Indochinese Clade. The Eastern Clade overlaps with the Northern Clade in the north, the Indochinese Clade in the middle, and the Sundaic Clade in the south. The Burmes Clade shares most of its distribution area with the Southern Clade.

Analyses of the concatenated dataset as well as ASTRAL analysis inferring a species tree from single gene trees revealed a basal dichotomy in the family Nemacheilidae (Fig. 1,2,4). The one branch consists of the Eastern Clade as sister to the pair Indochinese plus Northern Clade. The second branch is made of the Sundaic Clade as sister to the pair Burmese and Southern Clade. All basal nodes and most of tip nodes are supported by high statistical supports in all analyses.

Divergence times

The ages of the genealogical events were determined using three fossil records and one geological event as calibration points (details of calibration under ‘Methods’). The results indicate that Nemacheilidae shared the last common ancestor with other families of Cobitoidea about 48.5 mya (Fig. 4). The first notable split that separated the family into two big branches took place around 39 mya (for details see Fig. 5).

Topologies of main clades derived from various analyses.

A) Topologies recovered by ML analyses of single gene trees, with values at branches indicating bootstrap supports. Topologies congruent to the concatenated data-derived as well as species tree is revealed by Cyt b, MYH6 and IRBP2. Three slightly different topologies were derived from RH, EGR3 and RAG1 B) Topology obtained from analyses of concatenate dataset, the values at the branches correspond to ML/MrBayes PP/BEAST PP supports, respectively. C) Topology inferred by ASTRAL-III using the unrooted ML single gene trees as input. The values at the branches correspond to support values/gCF/sCF.

Divergence time estimation: Ingroup of the maximum clade credibility (MCC) tree resulting from Bayesian divergence time analysis of concatenated dataset in BEAST 2.

The red star indicates internal calibration point, the values at the nodes represent posterior probabilities (PPs lower than 0.90 not shown), and the blue bars relevant 95% HPD. The time scale is in millions of years. Inset in top left corner shows complete tree: Nemacheilidae is highlighted in yellow, while the outgroup is not highlighted; all calibration points are indicated by red stars.

Biogeographic reconstructions

The biogeographic reconstruction of Nemacheilidae considered 11 geographic regions within Eurasia and northeast Africa. The results show that Nemacheilidae originated in the region of nowadays Indochina and expanded from there in multiple waves. More details are given in Fig. 6.

Reconstruction of the biogeographical history of Nemacheilidae.

The inset map in top left corner shows the defined biogeographical regions of Eurasia recognised in the analysis; below the map the colour – region code for the main tree. Grey colour indicates combination of two regions as ancestral (FH); black colour indicates ancestral ranges identified with likelihood < 10%. Main tree: Ancestral range reconstruction with use of DEC+J model based on the phylogeny derived from BEAST2. Pie charts at the nodes indicate relative likelihood of ancestral ranges of most recent common ancestors (MRCA), while the current distribution of taxa is indicated at the tips.

Discussion

The evolutionary history of Nemacheilidae (Fig. 7, 8, Table 1)

Early Eocene (56 – 48 mya) - the origin of Nemacheilidae (Fig. 8A)

The family Nemacheilidae shared the last common ancestor with the other Cypriniformes about 48.5 mya (Fig. 2), and our biogeographic analysis identified the region of nowadays Indochina as most likely place of origin of Nemacheilidae (Fig. 6). During the early Eocene, Southeast Asia experienced a warm and perhumid climate, characterised by both highlands and lowlands with numerous rivers (Morley, 2018); conditions under which stone loaches flourished.

Time tree of the evolutionary history of Nemacheilidae.

The figure presents the evolutionary history of Nemacheildae, highlighting the dating of key geological or climatic key paleo-events that might have significantly impacted their evolution. Orange bars represent exceptionally warm periods, blue bars denote cold periods. Boxes indicate specific geological or climatic event. For detailed explanations of these events and their impact on nemacheilid evolution refer to text. Climatic states from Westerhold et al., 2020.

Paleomaps visualising the geographical and chronological aspects of the main events during the evolutionary history of the freshwater fish family Nemacheilidae across Eurasia.

A red star marks the point of origin, arrows indicate dispersal events. Background maps were redrawn according to Blakey (2023), Dierke Atlas (2023), Wang (2004), Westerweel et al. (2020).

Middle to late Eocene (48 – 34 mya) – first range expansions and formation of the first major clades (Fig. 8B,C)

The first detectable cladogenetic event within the family occurred about 39 mya and separated a lineage formed by the Eastern, Northern and Indochinese Clades (Branch A in Fig. 5) from a lineage including the Sundaic, Burmese and Southern Clades (Branch B). Our biogeographic analysis of the dataset revealed Indochina as most parsimony area of the first cladogenetic event; and consequently as region of origin for branch A as well as branch B.

The Eastern Clade was the first of the six present major clades to form: it separated about 35.8 mya from the common ancestor of the Indochinese and Northern Clades, either in the region of Indochina or Vietnam-south China. Taking into consideration its present-day distribution (Fig. 3) we assume that the young

Eastern Clade spread mainly along the eastern coastline of Asia southwards to Borneo and northwards to northeast China. Geological data support this hypothesis: During the Eocene, Asia was much smaller than today and had a rather flat relief with old mountains mainly along its eastern coast due to the collision with the Pacific Plate since 190 mya (Seton et al., 2012; Wang, 2004), while the upfolding of east-west mountain ridges had just started. The generally warm and humid climate during Eocene (Morley, 2018; Westerhold et al., 2020) additionally promoted the range expansion of the Eastern Clade. It can be assumed that shortly after the range expansion the distribution of the Eastern Clade was more or less continuous and not separated into several disjunct areas as nowadays.

The ancestral Indochinese + Northern Clades also spread northwards, but their present-day distribution (Fig. 3) suggests that they took a route through Central Indochina and China. Most likely, this first geographic expansion led the ancestral Indochinese+Northern Clades northward to at least the Tibetan Protoplateau, which during Eocene epoch had its northern boundary in the region of today’s Tanggula mountains (Wang et al., 2008), that means not even half of its present north-south extension. The presence of the two protoclades on the early Tibetan Plateau is indicated by the presence of some of their most basal members in the Yangtze basin and Yunnan: the genus Homatula for the Indochinese Clade and several species of Triplophysa for the Northern Clade (Figs. 1, 5).

Oligocene (34 – 23 mya) – range expansions, extinctions and the formation of the remaining major clades (Fig. 8D,E)

The Eocene - Oligocene transition brought a period of significant global cooling (Hutchinson et al., 2021) as well as a change in the precipitation regime of Central Asia (Clift and Webb 2018; Ye et al., 2022) due to the ongoing uplift of the Himalayan chain (Ding et al., 2017). A substantial decrease of rainfall in Central Asia caused an aridification from 34 to 23 mya (Bosboom et al., 2014) extending across nearly all of Oligocene Asia from the western to the eastern coast (Wang, 2004). The distribution area of the Eastern Clade became fragmented during this period (Fig. 3).

The Central Asian arid zone also split the Indochinese – Northern protoclade into two parts that evolved into two separate major clades, the Indochinese Clade and the Northern Clade. Moreover, the distribution area of the Indochinese Clade was split and separated the southernmost taxa (Speonectes and Tuberoschistura) from the taxa presently inhabiting all of Indochina and southern China, including the middle Yangtze basin (Fig. 3). The absence of the Indochinese Clade from any waters north of the Tibetan Plateau suggests that its Oligocene refuge was located southeast of the Tibetan Protoplateau. In contrast, the early Northern Clade found its refuge further north on the Tibetan Protoplateau, the only region that can explain the further history and the present distribution of the clade. By late Oligocene, the climate in Central Asia returned to a warmer and moister state (Zachos et al., 2001) and the Pamir and western Kunlun Mountains (in present day SE Tajikistan, N Pakistan and W China) rapidly thrust upward as an expansion of the northern Tibetan Plateau (Zheng et al., 2015). These favourable conditions coincided with a major split within the Northern Clade at 20 mya (Fig. 5), when the proto-Barbatula clade colonised the Pamir and Kunlun Mountains, while the proto Triplophysa branch found its main distribution rather east of the Tibetan Plateau.

Different geological events drove the evolution of branch B: the common ancestor of the Burmese and Southern Clades split at 32.5 mya in Indochina from the Sundaic Clade because the former left Southeast Asia by colonising the West Burma Terrane, a microplate that was pushed northward by the Indian plate (Morley et al., 2021) and made temporary contact with western Sundaland (today’s western Malay Peninsula) around Eocene – early Oligocene (Westerweel et al., 2019). Around 30 mya, it made contact with northeast India (Morley et al., 2021), allowing fishes from the West Burma Terrane to colonise the Indian subcontinent. During the Oligocene, the Indian plate and the West Burma Terrane both collided with mainland Asia. The Indian plate presently forms the Indian subcontinent, the West Burma Terrane now forms western Myanmar (Westerweel et al., 2020). Their parallel collision with mainland Asia caused the uplift of the north-south directed Indo-Burmese mountain range (Morley et al., 2020), separating the aquatic fauna of the Brahmaputra basin on the Indian side, from the Chindwin-Irrawaddy basin on the Burmese side, thus separating the Burmese Clade from the Southern Clade.

Miocene (23 – 5.3 mya) – rapid diversification (Fig. 8F,G)

The Miocene is generally an epoch of cooling, with temperatures decreasing from about 5.5° C above present conditions to conditions similar to present (Zachos et al., 2001). But temperatures also underwent strong fluctuations, leading to exceptionally warm periods like the Middle Miocene Climatic Optimum (17 – 14.5 mya) as well as to drastic cooling events like the Middle Miocene Climatic transition (14 mya) (Morley, 2018). Additionally, the Asian monsoon regime strengthened from the early Miocene (Clift and Webb, 2018), increasing precipitation and altering the distribution of climatic regions across Asia (Sun and Wang, 2005). Moreover, during the Miocene the Himalayan Mountains experienced their main uplift (Ding et al., 2017) and the upfolding area of Central Asia mountain ranges expanded northwards, giving rise to the Tian Shan and Altai ranges (Nissen et al., 2009; Zheng et al., 2015; Liu et al., 2023).

During this complex geological and climatic scenario, the family Nemacheilidae underwent a massive diversification. In our phylogenetic hypothesis, the family comprised 16 lineages at the beginning of Miocene. These lineages diversified during Miocene into 177 lineages. The radiation occurred across all six major clades, albeit to a lesser extent in the Eastern Clade. Interestingly, it manifested similarly in the region of the Indian monsoon and the East Asian monsoon. In both regions, major radiations occurred during the Middle Miocene Climatic Optimum, including the radiation of the Barbatula lineage in the Northern Clade, of the Indochinese Clade, the Sundaic Clade and of six branches in the Southern Clade. It appears that under increased temperature and precipitation freshwater fishes flourished, across all of Asia and under different monsoon systems.

One of the radiations that began during the Middle Miocene Climate Optimum involves Afronemacheilus, the sole nemacheilid genus found in Africa (Prokofiev and Golubtsov, 2013). Our results confirm that the ancestral Afronemacheilus separated from the Southern Clade about 18 - 17 mya. It could have utilised two potential pathways to Africa: one across the Arabian Peninsula and via the southern Danakil Isthmus, a broad landbridge that stretched from Djibouti to Aden from late Miocene to Pliocene (Bosworth et al., 2005; Steward and Murray, 2017). The second pathway would lead through the Levant and via Sinai into the Nile (‘Gomphotherium landbridge’ Harzhauser et al., 2007), and subsequently upstream into the Ethiopian highlands. Unfortunately, the nearly complete extinction of freshwater fishes from the Arabian Peninsula and from the Nile during Pleistocene aridification (Kotwicki and Salaimani, 2009; Leplongeon, 2021) complicates the reconstruction of its pathway to Africa.

In the Northern Clade, the Triplophysa lineage flourished during the Miocene. Its present wide distribution in southern and Central China suggests that it expanded from its Oligocene refuge on the eastern Part of the Tibetan Plateau in an eastern direction (Fig. 3). In contrast, the Barbatula branch expanded from its refuge on the western Tibetan Plateau northwards, traversing the uplifting Tian Shan and Altai ranges. This expansion can be traced from the geographic position of the most basal species (T. labiata -Tian Shan, B. altayensis – Altai), and then eastwards (B. compressirostris + B. dgebuadzei – west Mongolia, B. cf. toni – east Mongolia; tip species in Korea (B. indet Korea) and Japan (B. oreas).

During late Miocene, Barbatula also expanded westwards through Siberia and colonised Europe between 5.6 mya (separation of the European taxa from the Mongolian sister taxon) and 3.0 mya (oldest split within European Barbatula). During this cool and dry period, glaciations occurred on the northern hemisphere (Zachos et al., 2001), and in Siberia the mean temperature of the coldest month dropped below 0° for the first time during Cenozoic (Popova et al., 2012). Freezing of the northern margin of Eurasia prevented the Siberian rivers (which flow from south to north) from discharging their water into the Northern Ocean and produced extensive wetlands on the shallow relief of Siberia. These wetlands served as pathways for freshwater fishes from East Asia to European river systems during Pleistocene (Kalous et al., 2012), and we assume that the same mechanism brought Barbatula to Europe during Pliocene.

Pliocene and Pleistocene (5.3 – 0.01 mya) - Ice ages

During Pliocene and Pleistocene global cooling continued until massive glaciations repeatedly covered the north of Eurasia (de Schepper et al., 2014). However, the Pliocene also brought a much stronger seasonality to Eurasia, especially in the form of cold winters (Shen et al., 2018). Despite these significant changes in climatic conditions, the evolution of Nemacheilidae seems not to have experienced any dramatic negative impact, but reveals mainly local speciation events (e.g. the genus Barbatula colonised Eastern Siberia, Japan and the Korean Peninsula and the European Barbatula diversified geographically). We know from more detailed studies on Nemacheilidae (Bohlen et al., 2020a,b; Šlechtová et al., 2021) that during Pliocene and Pleistocene also geographic shifts as well as expansions and disappearances occurred, but since these happened on rather local scale they were too insignificant to reflect in the present intercontinental dataset.

Taxonomic implications

Besides its conclusions on biogeography and evolutionary history of the Eurasian freshwater fauna, the present dataset offers insights into the taxonomy of Nemacheilidae. Among the 37 analysed genera, 11 are represented by a single species in the dataset; 18 of the genera with several species in the dataset appear as monophyletic units in our analysis. However, seven genera (Mustura, Paraschistura, Physoschistura, Oxynoemacheilus, Schistura, Sectoria, Triplophysa) are identified as para- or polyphyletic based on our results. The most scattered genus is Schistura, which is listed in three of the six major clades and in a total of about 20 different genetic lineages.

In fact, this is nothing new. The ‘polyphyly’ of several larger genera, particularly Schistura, has been recognised for a long time: already Kottelat (1990) referred to Schistura and Nemacheilus as ‘wastebasket names’. These are not really cases of polyphyly; it is simply that a large number of new species have been described in the last 40 years and, not knowing in which genus to place them, authors simply ’dumped’ them in a few catch-all genera, waiting for more comprehensive studies and knowing well that the used genus names are mere labels. In recent years, more detailed morphological studies have resolved the relations of some groups of species within these catch-all genera (e.g. Conway and Kottelat, 2023; Dvořák et al., 2023; Kottelat, 2018, 2019).Moreover, every genetic study that included more than five species of Schistura resulted in polyphyly of the genus (e.g. Chen et al., 2019; Min et al., 2023; Sember et al., 2015; Sgouros et al., 2019; Tang et al., 2006). A recent detailed phylogenetic investigation of the genus Nemacheilus (Šlechtová et al., 2021) found the genus to be polyphyletic, but it became monophyletic after identifying five species that were falsely placed into Nemacheilus (these species are labelled ‘Nemacheilus’ in the present study). Similarly, revisionary studies of the genera Paracanthocobitis, Troglonectes, Micronemacheilus, Paranemachilus and Oreonectes, also allowed to resolve their intrarelationships and to redefine them into monophyletic taxa (Singer and Page, 2015; Luo et al., 2023). The main obstacles to a comprehensive taxonomic revision of Nemacheilidae is its vast size, a fast increasing number of new species, and the difficulty to access material from critical areas because of international and local political issues. We believe that results of the present study will facilitate future research on the taxonomic understanding of Nemacheilidae.

Another aspect of Nemacheilidae taxonomy is the existence of numerous undescribed species. Among the 279 species in our dataset, 36 (13%) were either clearly undescribed (‘Genus sp.’) or named with reservations (‘Genus’ cf. xxx’). This ratio is an underestimate, since, in general, undescribed species were avoided when composing the dataset. However, the existence of a vast amount of undescribed species within Nemacheilidae is illustrated by the fact that 430 of the 838 valid species of Nemacheilidae known in November 2024 (51%) have been described in the last 25 years (pers. obs.; Kottelat, 2012, 2013, updated).

Conclusions

- We reconstruct the evolutionary history of the Eurasian freshwater fish family Nemacheilidae during most of Cenozoic era. The extensive dataset (471 specimens, 279 species, six genes) from across Eurasia and the low dispersal capacity of the model resulted in high resolution of the phylogenetical and biogeographical analysis.

- Molecular phylogeny uncovered six major clades within the family. Dating of cladogenetic events and ancestral range estimation traced the origin of Nemacheilidae to Indochina around 48 million years ago.

- Major expansions were facilitated by tectonic connections, favourable climatic conditions, and orogenic processes. Conversely, aridification emerged as the primary cause of extinction events.

- Our study represents the first comprehensive and holistic reconstruction of the evolution of Eurasian freshwater biodiversity on a continental scale and through deep geological time.

- Our analyses identified numerous polyphyletic or paraphyletic taxa, highlighting the need for a thorough taxonomic revision within the family Nemacheilidae.

Methods

Sampling and Laboratory procedures

471 samples of more than 250 species from 37 genera of the family Nemacheilidae were examined, including the set of sequences of 364 specimens analysed in our laboratory and sequences of 107 specimens obtained from GenBank. Specimens used for the present study were fixed and stored in 96% ethanol. For more details about samples, sampling sites and GenBank accession numbers see Supplementary material Table S1.

Total genomic DNA was isolated using the DNeasy Blood and Tissue kit (Qiagen) following the manufacturer’s instructions. One mitochondrial gene (cytochrome b) and five nuclear genes (RAG1, IRBP2, MYH6, RH1 and EGR3) were selected. For list of primers see Supplementary material Table S2, details about PCRs settings are described in Chen et al., 2003; Chen et al., 2008 and Dvořák et al., 2022.

Sequencing of the PCR products was performed in Laboratory of Fish Genetics on ABI Prism 3130 GA or via sequencing service in Macrogen Europe BV (Amsterdam, Netherlands).

Phylogenetic analyses

Chromatograms were checked and assembled in the SeqMan II module of the DNA Star software package (LASERGENE). Single gene alignments were done in BioEdit (Hall, 1999) with use of ClustalW (Larkin et al., 2007) multiple alignment algorithm. The datasets were concatenated in PhyloSuite v1.2.2 (Zhang et al., 2020).

The best-fit substitution models and partitioning schemes (Supplementary material Table S3) were estimated using Partition Finder 2 (Lanfear et al., 2016) implemented in PhyloSuite 1.2.2 based on the corrected Akaike Information Criterion (AICc).

The 471 specimen phylogenetic trees were inferred from six loci concatenated dataset using the maximum likelihood (ML) and Bayesian inference (BI). The ML analyses were performed using IQ-TREE (Nguyen et al., 2015) implemented in PhyloSuite. The best-fit evolutionary model for each codon partition was automatically determined by ModelFinder (Kalyaanamoorthy, et al., 2017). The node support values were obtained with 5000 ultrafast bootstrap replicates (UFBoot) (Hoang et al., 2018). For the BI analyses we used MrBayes 3.2.7a (Ronquist and Huelsenbeck, 2003) on CIPRES Science Gateway (Miller et al., 2010). The datasets were partitioned into genes and codon positions and analyses were performed in two parallel runs of 10-20 million generations with 8 Metropolis Coupled Markov Chains Monte Carlo (MCMCMC). The relative burnin of 25% was used and from the remaining trees a 50% majority rule consensus trees were constructed.

For reconstructing of the species tree we used the Accurate Species TRee ALgorithm (ASTRAL III) (Zhang et al., 2018). For ASTRAL we used unrooted single gene ML trees reconstructed in IQ-TREE implemented in PhyloSuite. We have used IQ-TREE 2 (Minh at al., 2020) to calculate gene concordance factor (gCF, indicating the percentage of decisive gene trees containing particular branch) and site concordance factor (sCF, indicating the percentage of decisive alignment sites that support a branch in the reference tree) for every branch in the ASTRAL species tree. Distribution maps in Fig. 2 were designed using HYDROSHEDS (Lehner and Grill 2013). All original trees are available from authors upon request.

Divergence time estimations and ancestral range reconstruction

The ages of cladogenetic events were estimated in BEAST 2.6.4 (Bouckaert et al., 2014) via CIPRES Science Gateway with use of four calibration points. The first calibration point is based on the oldest known fossil of the family Catostomidae, Wilsonium brevipinne, from early Eocene (Liu, 2021), approximately 56-48 my old. The second calibration point is derived from the only known nemacheilid fossil record, Triplophysa opinata from Kyrgyzstan from middle-upper Miocene (16.0 to 5.3 mya) (Böhme and Ilg, 2003; Prokofiev, 2007). For some time, the Miocene fossil species Nemachilus tener from Central European was considered a member of Nemacheilidae. However, subsequent research has cast doubt on placement of this species within loaches (Obrhelova, 1967), leaving Triplophysa opinata as the only known fossil of Nemacheilidae. Third calibration point is based on the oldest known fossil of the genus Cobitis, C. naningensis, from early to middle Oligocene in southern China, 34-28 my old (Chen et al., 2015). The fourth calibration point, based on the river history of the southern Korean Peninsula, dates the separation between the species Cobitis tetralineata and C. lutheri to 2.5-3.5 my (Kwan et al., 2014). For the fossil-based calibration points we have used a fossil calibration prior implemented in CladeAge (Matschiner et al., 2017), the fourth calibration point was set to a uniform prior (2.5 – 3.5). The calibration points are depicted in Figs. 2 and 5.

For the final analyses, the partitions were unlinked and assigned the estimated evolutionary models. We used relaxed lognormal molecular clock and birth-death prior. Given the large and complex nature of our dataset, and in light of several preliminary analyses where MCMC chains did not provide satisfactory ESSs even after 1 billion iterations, we made the decision to streamline the dataset by retaining only one specimen per species. To enhance the performance of the analysis, we also implemented parallel tempering using the CoupledMCMC package (Müller and Bouckaert, 2019, 2020). The analysis was configured with four parallel chains of 6×108 generations, with resampling every 1000. Tree and parameter sampling intervals were set to 60.000. The effective sampling sizes (ESS) for all parameters were assessed in Tracer 1.7.1 (Rambaut et al., 2018). A maximum clade credibility (MCC) tree was built in TreeAnnotator 2.6.0 (Rambaut and Drummond, 2010) after discarding the first 25% of trees. The final trees were visualised in FigTree 1.4.4 (Rambaut, 2019). ML and BI were performed with both full as well as the reduced datasets.

The biogeographical analysis was conducted using the BioGeoBEARS package (Matzke, 2013) implemented in RASP 4.0 (Reconstruct Ancestral Stage in Phylogenies, Yu et al., 2015). We set 11 biogeographic regions according to drainage areas separated by major mountain ridges (Ural, Caucasus. Himalayas, Tenasserim Ridge, Indo-Burmese Range, Annamite Cordilliera, Balkan, Hindukush) or saltwater straits (Red Sea) (see Fig. 4A). The selected biogeographic regions are mostly congruent with the classical subregions of the Palaearctic (European, Siberian, East Asia and Mediterranean) and the Oriental (Indian, Malay and Chinese). For the analysis, we utilized the final MCC tree and 100 post burnin trees from the previous BEAST analysis. A maximum of 2-unit areas was allowed to be present in every reconstructed ancestral range. The comparison of six biogeographic models (DEC, DEC+J, DIVALIKE, DIVALIKE+J, BAYAREALIKE, BAYAREALIKE+J) suggested that the DEC+J is the best-suited model for the given dataset (as indicated by AICc weight value, Supplementary material Table 4). Likelihood Ratio Test (LRT) p-values, comparing models with and without the J factor, indicated that the addition of the J factor significantly influences the model likelihood. Considering recent criticisms of DEC and DEC+J (Ree and Sanmartin, 2018), we opted to conduct the analysis not only with DEC+J but also with other +J models. Comparison of the results from all applied models revealed only minimal differences.